---
title: "P8105 HW 2"
output: github_document
date: "2026-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Pull packages
```{r}
library(tidyverse)
```

### Problem 1

Read and clean data, retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert Yes and No to logical variable.

```{r}

transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
           na = "") |>
           janitor::clean_names() |>
  select(line:entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE),
    entry = as.logical(entry),
    vending = case_match(
      vending,
      "YES" ~ TRUE,
      "NO" ~ FALSE),
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
    )

# entry and vending have YES and NO values

# using str(transit_df) in console, found that the route columns variable types don't match, will need the mutate

```
line represents the physical subway line the subway route runs through, such as the West End, or Jerome in the Bronx. station_name indicates the name of a single station on the line. Each route column indicates indicates what route is served, such as the Q train, or 4/5/6 trains. If the station services less than 11 routes, the value is NA. station_latitude / longitude indicates the location of the subway station. entrance_type indicates the station's entrance type. Entry indicates whether or not the entrance type is also a point of entry. If false, it is an exit. Vending indicates if there is a MTA card vending machine at the entrance and ada notes whether or not it is ADA (Americans with Disabilities Act) compliant. So far, I have told `read_csv` to treat blank values as NA, cleaned the column names to be all lower case and standardized with `janitor::clean_names()`, selected the columns of interest with `select()`, and used the `mutate()` function to transform the character variables in entry and vending to logical variables. route8, 9, 10, and 11 column types were converted to characters since they are subway route names.

```{r}
# number of distinct stations 

transit_df |>
  distinct(line, station_name) |>
  nrow()

# number of stations ADA compliant

transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

# proportion of entrances/ exits without vending allow entrance

transit_df |>
  filter(vending == FALSE) |>
  pull(entry) |>
  mean()

```
There are 465 distinct stations. 84 stations are ada compliant. The proportion of station entrances/ exits without vending allow entrance is 0.3770492. 

Reformat data so route number and route name are distinct variables
```{r}

transit_long = 
  transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )

```

Distinct stations serve the A train. Of those, ADA compliant?
```{r}

transit_long |> 
  filter(route_name == "A") |>
  distinct(line, station_name, route_name) |>
  nrow()

# ADA compliant

transit_long |>
  filter(route_name == "A") |>
  filter(ada == TRUE) |>
  distinct(line, station_name, route_name) |>
  nrow()

```
60 distinct stations serve the A train. Only 17 of these 60 stations are ADA compliant.

### Problem 2

```{r}
trash_df = 
  readxl:: read_excel(
  "./data/202509 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel",
  skip = 1, #omit unnecessary rows
  range = "A2:N710") |>
  janitor::clean_names() |> #cleaning names
  filter(!is.na(dumpster)) |> #remove last NA column using dumpster
  mutate(
    sports_balls = as.integer(round(sports_balls, digits=0)), #round balls to nearest integer
    wheel_type = "mr_trash",
    year = as.integer(year)
  )

prof_df = 
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel",
  skip = 1, 
  range = "A2:M135") |>
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> #no sports balls
  mutate(
    wheel_type = "professor_trash"
  )

gwynnda_df = 
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
  sheet = "Gwynns Falls Trash Wheel",
  skip = 1, 
  range = "A2:L352") |>
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |>
  mutate(
    wheel_type = "gwynnda"
  )

trash_tidy = 
  bind_rows(trash_df, prof_df, gwynnda_df) |>
  relocate(wheel_type) |>
  mutate(
    month = str_to_lower(month)
  )

```

There are `r nrow(trash_tidy)` total observations in the new tidy dataset. Of these `r nrow(trash_tidy)` observations, `r filter(trash_tidy, wheel_type == "mr_trash") |> nrow()` are from Mr. Trash Wheel, `r filter(trash_tidy, wheel_type == "professor_trash") |> nrow()` are from Professor Trash Wheel, and `r filter(trash_tidy, wheel_type == "gwynnda") |> nrow()` are from the Gwynns Falls Trash Wheel. There are `r trash_tidy |> distinct(dumpster, wheel_type) |> nrow()` distinct dumpsters in total. The column month will output all 12 months of the year: `r trash_tidy |> pull(month) |> unique() |> knitr::combine_words()`. The year column contains years between `r trash_tidy |> pull(year) |> min()` and `r trash_tidy |> pull(year) |> max()`  weight_tons lists the total weight of trash collected by the trash wheel at the given observation. Variables such cigarette_butts, bottles, sports_balls, and wrappers count the the amount of each trash type collected by the trash wheel at the time of observation. The total weight of trash collected by Professor Trash Wheel was `r trash_tidy |> filter(wheel_type == "professor_trash") |> pull(weight_tons) |> sum()` tons. In June of 2022, Gwynnda collected `r trash_tidy |> filter(wheel_type == "gwynnda", month == "june", year == 2022) |> pull(cigarette_butts) |> sum() |> format(scientific = FALSE, big.mark = ",")` cigarette butts.