---
title: "P8105 HW 2"
output: github_document
date: "2026-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Pull packages
```{r}
library(tidyverse)
```

### Problem 1

Read and clean data, retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert Yes and No to logical variable.

```{r}

transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
           na = "") |>
           janitor::clean_names() |>
  select(line:entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE),
    entry = as.logical(entry),
    vending = case_match(
      vending,
      "YES" ~ TRUE,
      "NO" ~ FALSE),
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
    )

# entry and vending have YES and NO values

# using str(transit_df) in console, found that the route columns variable types don't match, will need the mutate

```
line represents the physical subway line the subway route runs through, such as the West End, or Jerome in the Bronx. station_name indicates the name of a single station on the line. Each route column indicates indicates what route is served, such as the Q train, or 4/5/6 trains. If the station services less than 11 routes, the value is NA. station_latitude / longitude indicates the location of the subway station. entrance_type indicates the station's entrance type. Entry indicates whether or not the entrance type is also a point of entry. If false, it is an exit. Vending indicates if there is a MTA card vending machine at the entrance and ada notes whether or not it is ADA (Americans with Disabilities Act) compliant. So far, I have told `read_csv` to treat blank values as NA, cleaned the column names to be all lower case and standardized with `janitor::clean_names()`, selected the columns of interest with `select()`, and used the `mutate()` function to transform the character variables in entry and vending to logical variables. route8, 9, 10, and 11 column types were converted to characters since they are subway route names.

```{r}
# number of distinct stations 

transit_df |>
  distinct(line, station_name) |>
  nrow()

# number of stations ADA compliant

transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

# proportion of entrances/ exits without vending allow entrance

transit_df |>
  filter(vending == FALSE) |>
  pull(entry) |>
  mean()

```
There are 465 distinct stations. 84 stations are ada compliant. The proportion of station entrances/ exits without vending allow entrance is 0.3770492. 

Reformat data so route number and route name are distinct variables
```{r}

transit_long = 
  transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )

```

Distinct stations serve the A train. Of those, ADA compliant?
```{r}

transit_long |> 
  filter(route_name == "A") |>
  distinct(line, station_name, route_name) |>
  nrow()

# ADA compliant

transit_long |>
  filter(route_name == "A") |>
  filter(ada == TRUE) |>
  distinct(line, station_name, route_name) |>
  nrow()

```
60 distinct stations serve the A train. Only 17 of these 60 stations are ADA compliant.

### Problem 2

```{r}
trash_df = 
  readxl:: read_excel(
  "./data/202509 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel",
  skip = 1, #omit unnecessary rows
  range = "A2:N710") |>
  janitor::clean_names() |> #cleaning names
  filter(!is.na(dumpster)) |> #remove last NA column using dumpster
  mutate(
    sports_balls = as.integer(round(sports_balls, digits=0)), #round balls to nearest integer
    wheel_type = "mr_trash",
    year = as.integer(year)
  )

prof_df = 
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel",
  skip = 1, 
  range = "A2:M135") |>
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> #no sports balls
  mutate(
    wheel_type = "professor_trash"
  )

gwynnda_df = 
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
  sheet = "Gwynns Falls Trash Wheel",
  skip = 1, 
  range = "A2:L352") |>
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |>
  mutate(
    wheel_type = "gwynnda"
  )

trash_tidy = 
  bind_rows(trash_df, prof_df, gwynnda_df) |>
  relocate(wheel_type) |>
  mutate(
    month = str_to_lower(month)
  )

```

There are `r nrow(trash_tidy)` total observations in the new tidy dataset. Of these `r nrow(trash_tidy)` observations, `r filter(trash_tidy, wheel_type == "mr_trash") |> nrow()` are from Mr. Trash Wheel, `r filter(trash_tidy, wheel_type == "professor_trash") |> nrow()` are from Professor Trash Wheel, and `r filter(trash_tidy, wheel_type == "gwynnda") |> nrow()` are from the Gwynns Falls Trash Wheel. There are `r trash_tidy |> distinct(dumpster, wheel_type) |> nrow()` distinct dumpsters in total. The column month will output all 12 months of the year: `r trash_tidy |> pull(month) |> unique() |> knitr::combine_words()`. The year column contains years between `r trash_tidy |> pull(year) |> min()` and `r trash_tidy |> pull(year) |> max()`  weight_tons lists the total weight of trash collected by the trash wheel at the given observation. Variables such cigarette_butts, bottles, sports_balls, and wrappers count the the amount of each trash type collected by the trash wheel at the time of observation. The total weight of trash collected by Professor Trash Wheel was `r trash_tidy |> filter(wheel_type == "professor_trash") |> pull(weight_tons) |> sum()` tons. In June of 2022, Gwynnda collected `r trash_tidy |> filter(wheel_type == "gwynnda", month == "june", year == 2022) |> pull(cigarette_butts) |> sum() |> format(scientific = FALSE, big.mark = ",")` cigarette butts.

### Problem 3

Importing and cleaning the MCI baseline dataset
```{r}
mci_bsl = read.csv(
  "./data/MCI_baseline.csv",
  skip = 1,
  na = "."
  ) |>
  janitor::clean_names() |>
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"),
    apoe4 = as.logical(apoe4)
    )

mci_bsl_tidy =
  mci_bsl |>
  filter(
    current_age < age_at_onset |
    is.na(age_at_onset)) |>
  mutate(
    id = as.character(id))

```

To clean the imported baseline dataset, first, we have to skip the first row of the csv file usign `skip = 1`, since this row acts as a dictionary for the columns, and is not relevant to anaylsis. Second, "." in the csv files was instructed to be treated as NA values. The sex variable was tranformed to character variables "male" and "female" instead of integers. apoe4, which notes APOE4 carrier status, was changed to be a TRUE/FALSE logical variable. In order to filter for the exclusion criteria, we must exclude study individuals that report MCI before or at the same age as the current age of enrollment. So, the `filter` function was used to keep individuals whose age at enrollment was less than the age of MCI onset as well as those who did not have an age of MCI onset recorded. `r nrow(mci_bsl)` participants originally enrolled in the study. Of the `r nrow(mci_bsl)` participants, `r  nrow(mci_bsl) - nrow(mci_bsl_tidy)` were excluded from the study. From this remaining cohort, `r mci_bsl_tidy |> filter(!is.na(age_at_onset)) |> nrow()` participants eventually developed MCI. After exclusion, the average baseline age was `r mci_bsl_tidy |> pull(current_age) |> mean()` years and the proportion of women in the study that are APOE4 carriers is `r mci_bsl_tidy |> filter(sex == "female") |> pull(apoe4) |> mean() * 100`% of them.

Import, clean, and tidy longitudinal biomarker values
```{r}
amyloid_df = 
  read_csv(
    "./data/mci_amyloid.csv",
    skip = 1,
    na = c("NA", "Na")
    ) |>
  janitor::clean_names() |>
  mutate(
    study_id = as.character(study_id)
  ) |>
  rename(id = study_id)

```

Check whether some participants appear only in one dataset
```{r}
anti_join(mci_bsl_tidy, amyloid_df, by = "id")
anti_join(amyloid_df, mci_bsl_tidy, by = "id")
```

`r nrow(anti_join(mci_bsl_tidy, amyloid_df, by = "id"))` participants appear in the baseline but not in the follow up period. `r nrow(anti_join(amyloid_df, mci_bsl_tidy, by = "id"))` appear in the follow up period but not at baseline.

Joining datasets, ensuring that only participants who appear in both datasets are retained.
```{r}
mci_clean = 
  inner_join(mci_bsl_tidy, amyloid_df, by = "id")
```
The resulting combined data set has data on `r nrow(mci_clean)` participants. The resulting dataset includes baseline values while also adding biomarker data to each corresponding participant id. 

Downloading as a CSV
```{r}
write_csv(mci_clean, "mci_clean.csv")
```



