---
title: "P8105 HW 2"
output: github_document
date: "2026-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Pull packages
```{r}
library(tidyverse)
```

### Problem 1

Read and clean data, retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert Yes and No to logical variable.

```{r}

transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
           na = "") |>
           janitor::clean_names() |>
  select(line:entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE),
    entry = as.logical(entry),
    vending = case_match(
      vending,
      "YES" ~ TRUE,
      "NO" ~ FALSE),
    )

# entry and vending have YES and NO values

view(transit_df)
```
line represents the physical subway line the subway route runs through, such as the West End, or Jerome in the Bronx. station_name indicates the name of a single station on the on the line. Each route column indicates whether or not this station/ line services a particular route, such as the Q train, or 4/5/6 trains. station_latitude / longitude indicates the location of the subway station. Entrance indiocates the station's entrance type. Entry indicates whether or not the entrance type is also a point of entry. Vending indicates if there is a MTA card vending machine in the station and ada notes whether or not it is ADA (Americans with Disabilities Act) compliant. So far, I have set NA values to be be blank spaces, cleaned the column names to be all lower case and standardized with `janitor::clean_names()`, selected the columns of interest with `select()`, and used the `mutate()` function to transform the character variables in entry and vending to logical variables.

```{r}



```