---
title: "P8105 HW 2"
output: github_document
date: "2026-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Pull packages
```{r}
library(tidyverse)
```

### Problem 1

Read and clean data, retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert Yes and No to logical variable.

```{r}

transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
           na = "") |>
           janitor::clean_names() |>
  select(line:entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE),
    entry = as.logical(entry),
    vending = case_match(
      vending,
      "YES" ~ TRUE,
      "NO" ~ FALSE),
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
    )

# entry and vending have YES and NO values

# using str(transit_df) in console, found that the route columns variable types don't match, will need the mutate

```
line represents the physical subway line the subway route runs through, such as the West End, or Jerome in the Bronx. station_name indicates the name of a single station on the line. Each route column indicates indicates what route is served, such as the Q train, or 4/5/6 trains. If the station services less than 11 routes, the value is NA. station_latitude / longitude indicates the location of the subway station. entrance_type indicates the station's entrance type. Entry indicates whether or not the entrance type is also a point of entry. If false, it is an exit. Vending indicates if there is a MTA card vending machine at the entrance and ada notes whether or not it is ADA (Americans with Disabilities Act) compliant. So far, I have told `read_csv` to treat blank values as NA, cleaned the column names to be all lower case and standardized with `janitor::clean_names()`, selected the columns of interest with `select()`, and used the `mutate()` function to transform the character variables in entry and vending to logical variables. route8, 9, 10, and 11 column types were converted to characters since they are subway route names.

```{r}
# number of distinct stations 

transit_df |>
  distinct(line, station_name) |>
  nrow()

# number of stations ADA compliant

transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

# proportion of entrances/ exits without vending allow entrance

transit_df |>
  filter(vending == FALSE) |>
  pull(entry) |>
  mean()

```
There are 465 distinct stations. 84 stations are ada compliant. The proportion of station entrances/ exits without vending allow entrance is 0.3770492. 

Reformat data so route number and route name are distinct variables
```{r}

transit_long = 
  transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )

```

Distinct stations serve the A train. Of those, ADA compliant?
```{r}

transit_long |> 
  filter(route_name == "A") |>
  distinct(line, station_name, route_name) |>
  nrow()

# ADA compliant

transit_long |>
  filter(route_name == "A") |>
  filter(ada == TRUE) |>
  distinct(line, station_name, route_name) |>
  nrow()

```
60 distinct stations serve the A train. Only 17 of these 60 stations are ADA compliant.